---
title: "kriging"
author: "environment"
date: "11/7/2021"
output: html_document
---
# Upload needed packages and source 'processing' file
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sp) # spatial point data frames
library(gstat) # variogram estimation
library(tmap)
library(geosphere)
library(sf)
library(tidyverse)
library(magrittr)
library(arm)
#library(scico)
```

# combine the data from dem
```{r}
env <- st_read("env.shp")
env <- env[!is.na(env$mean_pm25),] #Remove rows with null columns
env <- env[!is.na(env$mean_temp),] 
env <- st_centroid(env) # Calculate centroids
# epsg_wgs84 <- 4326
# env <- env %>% st_transform(epsg_wgs84)
# ggplot(env) + geom_sf()


dem <- st_read("DemographicsSmoothed2.shp")
dem <- st_centroid(dem)
#https://stackoverflow.com/questions/49032217/inner-joining-two-sf-objects-by-non-sf-column

#join selected environment columns to demographics
env_sub <- as.data.frame(env)
env_sub <- env_sub[,c('GEOID20','mean_temp','mean_pm25')]
jn <- env_sub  %>% inner_join(as.data.frame(dem)) %>% st_sf(sf_column_name = 'geometry') 
jn %<>% st_sf(sf_column_name = 'geometry')

#Standardize the coordinate system used
epsg_wgs84 <- 4326
jn <- jn %>% st_transform(epsg_wgs84)
```
#Spherical and Gaussian Variogram functions

```{r}
spherical_variogram <- function (n, ps, r) function (h) {
  h <- h / r
  n + ps * ifelse(h < 1, 1.5 * h - .5 * h ^ 3, 1)
}

gaussian_variogram <- function (n, ps, r)
  function (h) n + ps * (1 - exp(-(h / r) ^ 2))


exponential_variogram <- function (n, ps, r)
  function (h) n + ps * (1 - exp(-(h / r)))

# solves `A * x = v` where `C = chol(A)` is the Cholesky factor:
chol_solve <- function (C, v) backsolve(C, backsolve(C, v, transpose = TRUE)) # the solution of triangular system.The result will be a vector if 'x' is a vector and a matrix if 'x' is a matrix.



#New Kriging Function
epsilon <- 1e-3
kriging_smooth_spherical <- function (formula, data, ...) {
  v <- variogram(formula, data)
  v_fit <- fit.variogram(v, vgm("Sph",...))
  v_fit$psill[1] <- max(v_fit$psill[1], epsilon)
  v_f <- spherical_variogram(v_fit$psill[1], v_fit$psill[2], v_fit$range[2])
  
  Sigma <- v_f(as.matrix(dist(st_coordinates(data)))) # semivariogram
  Sigma <- sum(v_fit$psill) - Sigma # prior variance
  tau2 <- v_fit$psill[1] # residual variance
  C <- chol(tau2 * diag(nrow(data)) + Sigma)
  y <- model.frame(formula, data)[, 1] # response
  x <- model.matrix(formula, data)
  # generalized least squares:
  xt <- backsolve(C, x, transpose = TRUE)
  beta <- coef(lm.fit(xt, backsolve(C, y, transpose = TRUE))) # prior mean
  names(beta) <- colnames(x)
  beta_se <- sqrt(diag(chol2inv(chol(crossprod(xt)))))
  
  chol_sigma <- chol(Sigma)
  Sigma_inv <- chol2inv(chol_sigma)
  C <- chol(Sigma_inv + diag(nrow(data)) / tau2)
  # posterior mean (smoother):
  mu <- drop(chol_solve(C, y / tau2 + Sigma_inv %*% x %*% beta))
  
  sinv_mu <- chol_solve(chol_sigma, mu - x %*% beta)
  krige <- function (new_data) { # prediction function
    D <- apply(coordinates(data), 1,
               function (coord) apply(st_coordinates(new_data), 1, dist2, coord))
    V <- sum(v_fit$psill) - v_f(D)
    
    t <- delete.response(terms(formula))
    xp <- model.matrix(t, model.frame(t, new_data))
    drop(xp %*% beta + V %*% sinv_mu)
  }
  
  list(smooth = mu, prior_coef = beta, prior_coef_se = beta_se,
       variogram = v_fit, krige = krige)
}

# 
# kriging_smooth_gaussian <- function (formula, data, ...) {
#   # formula <- mean_pm25 ~ 1
#   # data <- env
#   v <- variogram(formula, data)
#   v_fit <- fit.variogram(v, vgm("Gau"))
#   v_f <- gaussian_variogram(v_fit$psill[1], v_fit$psill[2], v_fit$range[2])
#   Sigma <- v_f(as.matrix(dist(matrix(unlist(data$geometry), ncol = 2, byrow = TRUE)))) # semivariogram
#   Sigma <- sum(v_fit$psill) - Sigma # prior variance
#   tau2 <- v_fit$psill[1] # residual variance
#   C <- chol(tau2 * diag(nrow(data)) + Sigma)
#   y <- model.frame(formula, data)[, 1] # response
#   x <- model.matrix(formula, data)
#   # generalized least squares:
#   beta <- coef(lm.fit(backsolve(C, x, transpose = TRUE),
#                       backsolve(C, y, transpose = TRUE))) # prior mean
# 
#   Sigma_inv <- chol2inv(chol(Sigma))
#   C <- chol(Sigma_inv + diag(nrow(data)) / tau2)
#   # posterior mean (smoother):
#   mu <- drop(chol_solve(C, y / tau2 + Sigma_inv %*% x %*% beta))
#   list(smooth = mu, prior_mean = beta)}
# 
# 
# 
# kriging_smooth_exponential <- function (formula, data, ...) {
#   formula <- mean_pm25 ~ 1 + Mdn_Inc_s + Mdn_Inc_r + MnrPrc_s + MnrPrc_r
#   data <- env
#   v <- variogram(formula, data)
#   v_fit <- fit.variogram(v, vgm("Exp"))
# 
#   v_f <- exponential_variogram(v_fit$psill[1], v_fit$psill[2], v_fit$range[2])
# 
#   Sigma <- v_f(as.matrix(dist(matrix(unlist(data$geometry), ncol = 2, byrow = TRUE)))) # semivariogram
#   Sigma <- sum(v_fit$psill) - Sigma # prior variance
#   tau2 <- v_fit$psill[1] # residual variance
#   C <- chol(tau2 * diag(nrow(data)) + Sigma)
#   y <- model.frame(formula, data)[, 1] # response
#   x <- model.matrix(formula, data)
#   # generalized least squares:
#   beta <- coef(lm.fit(backsolve(C, x, transpose = TRUE),
#                       backsolve(C, y, transpose = TRUE))) # prior mean
# 
#   Sigma_inv <- chol2inv(chol(Sigma))
#   C <- chol(Sigma_inv + diag(nrow(data)) / tau2)
#   # posterior mean (smoother):
#   mu <- drop(chol_solve(C, y / tau2 + Sigma_inv %*% x %*% beta))
#   list(smooth = mu, prior_mean = beta)}

```

# old kriging function
```{r}

# kriging_smooth_spherical <- function (formula, data, ...) {
#   # formula <- mean_pm25 ~ 1
#   # data <- env
#   v <- variogram(formula, data)
#   v_fit <- fit.variogram(v, vgm("Sph"))
#   v_f <- spherical_variogram(v_fit$psill[1], v_fit$psill[2], v_fit$range[2])
#   #Sigma <- v_f(as.matrix(dist(coordinates(data)))) # previous code
#   Sigma <- v_f(as.matrix(dist(matrix(unlist(data$geometry), ncol = 2, byrow = TRUE)))) # semivariogram
#   Sigma <- sum(v_fit$psill) - Sigma # prior variance
#   tau2 <- v_fit$psill[1] # residual variance
#   C <- chol(tau2 * diag(nrow(data)) + Sigma)
#   y <- model.frame(formula, data)[, 1] # response
#   x <- model.matrix(formula, data)
#   # generalized least squares:
#   beta <- coef(lm.fit(backsolve(C, x, transpose = TRUE),
#                       backsolve(C, y, transpose = TRUE))) # prior mean
# 
#   Sigma_inv <- chol2inv(chol(Sigma))
#   C <- chol(Sigma_inv + diag(nrow(data)) / tau2)
#   # posterior mean (smoother):
#   mu <- drop(chol_solve(C, y / tau2 + Sigma_inv %*% x %*% beta))
#   list(smooth = mu, prior_mean = beta)
# }
```

names(jn)
 [1] "GEOID20"   "mean_temp" "mean_pm25" "OBJECTI"   "STATEFP"   "COUNTYF"   "TRACTCE"  
 [8] "NAME20"    "NAMELSA"   "MTFCC20"   "FUNCSTA"   "ALAND20"   "AWATER2"   "INTPTLA"  
[15] "INTPTLO"   "Sh_STA_"   "Sh_STL_"   "Tre_Cnp"   "TotalPp"   "MinrPrc"   "PpDnsty"  
[22] "MedGrTx"   "Mdn_Inc"   "MnrPrc_t"  "MnrPrc_s"  "MnrPrc_r"  "MdGrTx_t"  "MdGrTx_s" 
[29] "MdGrTx_r"  "Mdn_Inc_t" "Mdn_Inc_s" "Mdn_Inc_r" "tree_t"    "tr_smth"   "tre_rsd"  
[36] "geometry"

# smooth mean_pm25
```{r}
#Calculate variograms for mean_pm25
formula <- mean_pm25 ~ 1 + Mdn_Inc_s  + Mdn_Inc_r +log(PpDnsty) + tr_smth + tre_rsd
dt <- jn
v <- variogram(formula, dt)
v_fit <- fit.variogram(v, vgm("Sph")) #'vgm':generate a variogram model or add to an existing model.
v_f <- spherical_variogram(v_fit$psill[1], v_fit$psill[2], v_fit$range[2]) #Create spherical variograms
#Plot the variograms
op <- par(mfrow = c(1, 2))
h <- seq(0, 8, length = 1000)

plot(v$dist, v$gamma,  pch = 19, col = "gray",
     xlab = "distance", ylab = "semivariogram") # main = "Minority Percentage"
lines(h, v_f(h))
abline(v = v_fit$range[2], col = "gray")
plot(h, sum(v_fit$psill) - v_f(h), type = "l",
     xlab = "distance", ylab = "covariogram",
     ylim = c(0, sum(v_fit$psill)))
points(0, sum(v_fit$psill), pch = 19)
abline(v = v_fit$range[2], col = "gray")
par(op)

# check smoothing:
ks <- kriging_smooth_spherical(formula, dt)
y <- jn$mean_pm25
op <- par(mfrow = c(1, 2))
plot(ks$smooth, y); abline(0, 1, col = "red")
plot(ks$smooth, type = "l", ylab = "y")
points(y, pch = 19, col = "gray")
abline(h = ks$prior_mean)
par(op)

rbind(ks$prior_coef, ks$prior_coef_se) #to see coef values

jn$mean_pm25_smooth <- ks$smooth
jn$mean_pm25_resid <- jn$mean_pm25 - jn$mean_pm25_smooth 
# 
# t <- st_read("Census2020_Tracts/Census2020_Tracts.shp")
# data <- as.data.frame(check)
# jn <- jn  %>% inner_join(as.data.frame(t)) %>% st_sf(sf_column_name = 'geometry') 

# pm25 <- ggplot(data = jn, aes(fill = mean_pm25)) +
#   geom_sf() +
#   labs(title="PM2.5") +
#   scale_fill_distiller(palette = "Greys",trans = "reverse") +
#   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black")) +theme_void()
# 
# s_pm25 <- ggplot(data = jn, aes(fill = mean_pm25_smooth)) +
#   geom_sf() +
#   labs(title="Smooth_PM2.5") +
#   scale_fill_distiller(palette = "Greys",trans = "reverse") +
#  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black")) +theme_void()
# library(ggpubr)
# pm25 <- ggarrange(pm25, s_pm25, ncol=2, common.legend = TRUE, legend = "right")
# pm25
```

# smooth mean_temp
```{r}
#Calculate variograms for mean_temp

formula <- mean_temp ~ 1 + Mdn_Inc_s  + Mdn_Inc_r +log(PpDnsty) + tr_smth + tre_rsd
dt <- jn

v <- variogram(formula , dt)
v_fit <- fit.variogram(v, vgm("Sph")) #'vgm':generate a variogram model or add to an existing model.
v_f <- spherical_variogram(v_fit$psill[1], v_fit$psill[2], v_fit$range[2]) #Create spherical variograms
#Plot the variograms
op <- par(mfrow = c(1, 2))
h <- seq(0, 8, length = 1000)

plot(v$dist, v$gamma,  pch = 19, col = "gray",
     xlab = "distance", ylab = "semivariogram") # main = "Minority Percentage"
lines(h, v_f(h))
abline(v = v_fit$range[2], col = "gray")
plot(h, sum(v_fit$psill) - v_f(h), type = "l",
     xlab = "distance", ylab = "covariogram",
     ylim = c(0, sum(v_fit$psill)))
points(0, sum(v_fit$psill), pch = 19)
abline(v = v_fit$range[2], col = "gray")
par(op)

# check smoothing
ks2 <- kriging_smooth_spherical(formula , dt)
y2 <- jn$mean_temp
op <- par(mfrow = c(1, 2))
plot(ks2$smooth, y2); abline(0, 1, col = "red")
plot(ks2$smooth, type = "l", ylab = "y")
points(y2, pch = 19, col = "gray")
abline(h = ks2$prior_mean)
par(op)

rbind(ks2$prior_coef, ks2$prior_coef_se) #to see coef values

jn$mean_temp_smooth <- ks2$smooth
jn$mean_temp_resid <- jn$mean_temp - jn$mean_temp_smooth 
```



# pointplot: before-and-after smooth
```{r}
#Add smoothing and residuals to datasets

# plot
g1 <- ggplot(jn) + geom_sf(aes(fill = mean_pm25, color = mean_pm25)) + 
  scale_fill_viridis_c(direction = -1)+
  scale_color_viridis_c(direction = -1)

g2 <- ggplot(jn) + geom_sf(aes(fill = mean_pm25_smooth, color = mean_pm25_smooth)) + 
  scale_fill_viridis_c(direction = -1)+
  scale_color_viridis_c(direction = -1)
library(ggpubr) #install.packages("ggpubr")
ggarrange(g1,g2,common.legend = TRUE)

g3 <- ggplot(jn) + geom_sf(aes(fill = mean_temp, color = mean_temp)) + 
  scale_fill_viridis_c(direction = -1)+
  scale_color_viridis_c(direction = -1)
  
g4 <- ggplot(jn) + geom_sf(aes(fill = mean_temp_smooth, color = mean_temp_smooth)) + 
  scale_fill_viridis_c(direction = -1)+
  scale_color_viridis_c(direction = -1)
ggarrange(g3,g4,common.legend = TRUE)


```

#  qlogis(Tre_Cnp) ~  
```{r}
# formula <- qlogis(Tre_Cnp) ~  Mdn_Inc_s  + Mdn_Inc_r +log(PpDnsty) + mean_pm25_smooth + mean_pm25_resid +mean_temp_smooth + mean_temp_resid
# dt <- jn

# v <- variogram(formula, dt) 
# v_fit <- fit.variogram(v, vgm("Sph"))
# v_f <- spherical_variogram(v_fit$psill[1], v_fit$psill[2], v_fit$range[2])
# 
# op <- par(mfrow = c(1, 2))
# h <- seq(0, 200, length = 1000) 
# 
# plot(v$dist, v$gamma,  pch = 19, col = "gray",
#      xlab = "distance", ylab = "semivariogram", main = "Tree Canopy")
# lines(h, v_f(h))
# abline(v = v_fit$range[2], col = "gray")
# 
# # check smoothing:
# ks <- kriging_smooth_spherical(formula, dt)
# y <- qlogis(jn$Tre_Cnp)
# op <- par(mfrow = c(1, 2))
# plot(ks$smooth, y); abline(0, 1, col = "red") 
# plot(ks$smooth, type = "l", ylab = "y")
# points(y, pch = 19, col = "gray") 
# abline(h = ks$prior_coef) 
# par(op)
# 
# rbind(ks$prior_coef, ks$prior_coef_se) #to see coef values
# 
# # add tree canopy smooth
# jn$mean_cnp_smooth <- ks$smooth
# jn$mean_cnp_resid <- jn$Tre_Cnp - jn$mean_cnp_smooth 
```

# output smooth data
```{r}

# send_shape <- jn
# st_write(send_shape, "Tree_dem_env.shp")
# check <- st_read("Tree_dem_env.shp", quiet = TRUE)
# t <- st_read("Census2020_Tracts/Census2020_Tracts.shp")


```

# polygan plot: before-and-after smooth

```{r}

check <- st_read("Tree_dem_env.shp", quiet = TRUE)
t <- st_read("Census2020_Tracts/Census2020_Tracts.shp")

#join selected environment columns to demographics
data <- as.data.frame(check)
data_sub <- data[,c('GEOID20','men_tmp','mn_tmp_s', "mn_pm25", "mn_pm25_s","tree_t","tre_rsd","tr_smth")]

# data_sub <- data %>% dplyr::select(- 'geometry')
jn <- data_sub  %>% inner_join(as.data.frame(t)) %>% st_sf(sf_column_name = 'geometry') 
jn %<>% st_sf(sf_column_name = 'geometry')

#Standardize the coordinate system used
epsg_wgs84 <- 4326
jn <- jn %>% st_transform(epsg_wgs84)
library(ggpubr)


# After-Smoothing Map & Kriging Predicted Map for temp
jn$temp_refit <- ks2$smooth

t$temp_refit[t$GEOID20 %in% jn$GEOID20] <- jn$temp_refit[jn$GEOID20 %in% t$GEOID20]
ggplot() + geom_sf(color = 'grey', data = t) + geom_sf(aes(fill = temp_refit), data=na.omit(t)) + 
  scale_fill_distiller(palette="Blues", trans = "reverse") + labs(title = 'Original Model for Mental Health')
grid <- st_centroid(t[is.na(t$temp_refit),]) 
pred_MH <-  krige(temp_refit ~ 1, jn, grid$geometry, model = v_fit) # Error in .local(formula, locations, ...) : not all sf::st_crs(locations) == sf::st_crs(newdata) are TRUE
t$temp_refit[is.na(t$temp_refit)] <- pred_MH$var1.pred
ggplot(data = t) + geom_sf(color = 'grey') + geom_sf(aes(fill = temp_refit)) + 
  scale_fill_distiller(palette="Blues", trans = "reverse") + labs(title = 'Refitted Model for Mental Health ')

# the code from Health group
# After-Smoothing Map & Kriging Predicted Map for Mental Health
newhealth$ks_mh_refit <- ks_mh_refit$smooth
tract7$ks_mh_refit[tract7$GEOID20 %in% newhealth$GEOID20] <- newhealth$ks_mh_refit[newhealth$GEOID20 %in% tract7$GEOID20]
ggplot() + geom_sf(color = 'grey', data = tract7) + geom_sf(aes(fill = ks_mh_refit), data=na.omit(tract7)) + 
  scale_fill_distiller(palette="Blues", trans = "reverse") + labs(title = 'Original Model for Mental Health')
grid <- st_centroid(tract7[is.na(tract7$ks_mh_refit),]) 
pred_MH <-  krige(ks_mh_refit ~1, newhealth, grid$geometry, model = v_fit)
tract7$ks_mh_refit[is.na(tract7$ks_mh_refit)] <- pred_MH$var1.pred
ggplot(data = tract7) + geom_sf(color = 'grey') + geom_sf(aes(fill = ks_mh_refit)) + 
  scale_fill_distiller(palette="Blues", trans = "reverse") + labs(title = 'Refitted Model for Mental Health ')

```

